Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-05-26 doc/invt.tex
simplify invt.tex

The schedule of reducing
the updating magnitude,
therefore, critically affects
the precision of the final bias potential,
and hence that of the PMF\cite{laio2005, bussi2006, poulain2006,
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007, min2007,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014,
barducci2008, dickson2011, dama2014}.
%
For the single-bin case, the optimal updating schedule
is a function of the
inverse time\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007,
morozov2007, zhou2008}
or
the inverse number of updating steps.
\note{
  About the origin of the inverse-time formula.
  Besides the Belardinelli paper,
  we can also find it in Eq. (6) of reference \cite{liang2007},
  if I am not too mistaken.
  But I find the paper is a bit hard to read.
  There, however, the formula is $t_0/t$ instead of $1/t$.
}
%
%The optimal schedule for metadynamics,
%however, is less studied.

In this study,
we present a method of computing
the optimal schedule
for an adaptive FDS method
of a general updating scheme,
including the single-bin and Gaussian one
used in the WL algorithm and metadynamics.
%
Our method maps the optimization problem to a mechanical one,
in which the schedule plays the role of the velocity of
a free particle with a position-dependent mass,
and the final error becomes the action.
%
In this way, the optimal schedule
can be deduced from the equation of motion
of the particle that minimizes the action.
%
The resulting optimal schedule
depends on the updating window function
through the mass function.
%
While the optimal schedule in the single-bin case
is given by the known inverse-time formula,
that for Gaussian case is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
%In either case, however,
Ideally,
the optimal initial updating magnitude
can be set to
half of the previous equilibrium value,
which amounts to a shift of the origin of time
in the single-bin case.
%
Further, we will show that
the single-bin updating scheme
is optimal for asymptotic convergence,
and it can be generalized
to a class of asymptotically optimal
bandpass updating schemes.
%
The latter, however,
allow one to selectively flatten only
a few long-range histogram modes,
and to avoid introducing local noise
into the bias potential during the updating process.
%
The optimal schedule of bandpass updating schemes
is also the above origin-shifted inverse-time schedule.
%
Interestingly, several previous FDS algorithms\cite{
  langfeld2012, pellegrini2014,
  neuhaus2006, *neuhaus2007}

@@

Consider the problem of computing
the distribution, $p^*_i$,
along a discrete quantity, $i$,
for a given system.
%
%The discreteness of $i$ is natural
%in discrete models.
%
For example, $i$ can be the energy, $E$,
in a lattice spin model or the temperature index
in a simulated tempering simulation\cite{
marinari1992, *lyubartsev1992}.
%
For a continuous quantity, $z$,
which is often the case in a (classical) molecular system,
we can discretize $z$
such that the integer $i$ represents
the index of a small interval, or a bin,
$(z, z + \Delta z)$.
%
The distribution is normalized as
$\sum_{i = 1}^n p^*_i = 1$.

%For a large system at temperatures of interest,
%the distribution, $p^*_i$, often tends to
%be localized,
%%
%and to explore the global shape
%of the PMF, $-\ln p^*_i$,
%it is often advantageous to perform
%biased sampling that targets
%a wider and smoother distribution, $p_i$.
%%
%%Here, we refer to simulations that target
%%a flat or nearly flat distribution
%%as entropic or multicanonical sampling.
%
%To do so, we introduce a bias potential, $V_i$,
If the distribution, $p^*_i$, is multimodal,
it is often advantageous to perform
biased sampling that targets
a smoother distribution, $p_i$,
by introducing a bias potential, $V_i$,
such that the biased equilibrium distribution is
%
\begin{equation}
  \pi_i
  %=
  %C_i \, p^*_i \, e^{-V_i}
  =
  \frac{                p^*_i \, e^{-V_i} }
       { \sum_{j = 1}^n p^*_j \, e^{-V_j} }
  \propto
  p^*_i \, e^{-V_i}
  .
  \label{eq:pi_p_V}
\end{equation}
%
%where the proportionality constant, $C_i$,
%is found from the normalization condition,
%$\sum_{i = 1}^n \pi_i = 1$,
%in the second step.
%
%In particular,
%to make $\pi_i$ flat,
%the bias potential $V_i$
%must coincide with $\ln p^*_i$,
%up to an additive constant.
%%
%In practice, $V_i$ may be non-optimal or contain error,
%and thus $\pi_i$ need not
%coincide with the intended distribution $p_i$.

We will broadly define
a flat-distribution-sampling (FDS) simulation
as a biased simulation
targeting a flat $p_i$ (i.e. $p_i \equiv 1/n$)
or nearly flat one\cite{
dayal2004, *trebst2004, *zhang2007, barducci2008, singh2011}.
%
For the biased distribution, $\pi_i$,
to converge to the desired $p_i$,
the bias potential must satisfy
%
\begin{equation}
  V_i \to \ln p^*_i - \ln p_i + \mathrm{const.}
  .
  \label{eq:Vi_target}
\end{equation}
%
%according to Eq. \eqref{eq:pi_p_V}.
%which allows us to deduce the PMF, $-\ln p^*_i$,
%directly from the bias potential.
%
This relation thus sets a practical goal for the convergence of FDS
methods---the bias potential must be adjusted to make the
observed distribution match the intended one.
%%
Once this is done,
the PMF, $-\ln p^*_i$, can be naturally
deduced from the bias potential. % $V_i$.
\note{
  For example, in the context of simulated tempering (Zhang and Ma, 2007),
  $p^*_i$ is replaced by $\ln Z_i$,
  $p_i$ is the weight $\zeta_i$,
  and the bias potential, $V_i$, is $\ln (\tilde Z_i / \zeta_i)$.
  Upon convergence, we have $\ln \tilde Z_i \to \ln Z_i$
  up to a constant shift.
}

In terms of how one adjusts the bias potential,
there are roughly two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
uses a fixed bias potential, $V_i$,
(estimated a priori)
during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{
berg1992, *lee1993}
and simulated tempering\cite{
marinari1992, *lyubartsev1992}.
%with $i$ being the index of energy
%and that of temperature, respectively.
%
After a sufficiently long run, the bias potential, $V_i$,
can often be improved
with the normalized histogram, $H_i$,
accumulated from the simulation as
%
%Using $H_i$ as an estimate of $\pi_i$ in
%Eq. \eqref{eq:pi_p_V}
%allows us to correct $V_i$ as
%
\begin{equation}
  V^\mathrm{corr}_i
  =
  V_i
  +
  \ln \frac{ H_i }
           { p_i }.
  \label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    H_i \approx \pi_i
    &\propto p^*_i \, e^{-V_i},
    \\
    p_i
    &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
%However, for an unknown system,
%it is often difficult
%to guess a sufficiently accurate
%$V_i$ with which to start.
%%
%Thus,
%it can take many iterative rounds of simulations
%to achieve the desired flat distribution.
%%
%%Another difficulty of this approach is that
%The effectiveness of this approach
%depends on the precision of the histogram, $H_i$,
%which in turn depends on the simulation length
%that must be decided in advance.
However, this formula often requires a long
accumulation period for a precise histogram,
and thus disallows
continuous improvement of the bias potential
%(hence that of the PMF)
as simulation lengthens.

The second type of methods,
including the WL algorithm and metadynamics,
addresses the above shortcomings
by continuously and incrementally updating the bias potential.
%For this reason, the second type of methods,
%referred to as the adaptive FDS methods,
%were invented.
%
These methods, referred to as the adaptive FDS methods,
will be discussed below.
%%
%%These methods incrementally update the bias potential
%%and can approximately sample the target distribution.
%%
%However,
%the on-the-fly updates break the microscopic reversibility
%of the underlying sampling mechanism.
%Further, failure to properly reduce the magnitude of
%the histogram-based increments
%would lead to roughness or errors in the final estimate of the PMF.
%%
%%Thus, for a conservative practitioner,
%%the E-FDS methods appear more appealing
%%once a sufficiently accurate bias potential is available.
%%
%To minimize this effect,
%many adaptive methods adopt a schedule to
%reduce the updating magnitude
%over the course of simulation\cite{
%marsili2006,
%belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
%liang2007,
%barducci2008}.
%%
%%Thus, in late stages of a long simulation,
%%the updating becomes negligible and
%%the adaptive simulation tends to
%%the corresponding equilibrium one.
%%
%%if the bias potential were converged to the exact value,
%%this adaptive FDS tends to the ideal equilibrium counterpart,
%%particular equilibrium simulation can be thought as
%%the optimal asymptotic limit of the adaptive counterpart,
%%
%For example, one can show that
%if the magnitude is controlled by
%the optimal inverse-time prescription\cite{
%belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
%liang2007,
%morozov2007, zhou2008,
%komura2012, *caparica2012, *caparica2014},
%the WL algorithm is equivalent to the ideal equilibrium counterpart
%in improving the precision of the bias potential,
%thus allowing for maximum efficiency
%(cf. Appendix \ref{sec:equilerr}).
%%the bias potential in the WL algorithm
%%will evolve similar to Eq. \eqref{eq:vcorr_equil}
%%with the precision of the histogram as simulation lengthens.

@@

%We will mainly focus on the adaptive methods below.
%
In the WL algorithm\cite{wang2001, *wang2001pre},
the bias potential, $V_i(t)$, is updated
at each sampling step $t$ as
%
\begin{equation}
  V_i(t+1)
  =
  V_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { p_{i(t)} }
  ,
\label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
$\delta_{i, \, i(t)}$ is the Kronecker delta function,
which is $1$ when $i = i(t)$ or $0$ otherwise,
and $\alpha(t)$ is the updating magnitude.
%
We refer to this scheme as the \emph{single-bin scheme},
for it applies only to the bias potential
at the current bin, $i(t)$.
%
In a more general \emph{multiple-bin scheme},
including the Gaussian updating scheme
used in metadynamics,
several neighboring bins are updated:
%
\begin{equation}
  V_i(t+1)
  =
  V_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) }
       { p_{i(t)} },
  \label{eq:mbin_update}
\end{equation}
%
where the elements, $w_{ij}$'s, are a set of coefficients satisfying
the restrictions to be discussed
in Sec. \ref{sec:updating-matrix}.
%
Note that if the updating occurs
only every few sampling steps,
then $t$ means the number of updating steps.

@@

%For the single-bin scheme
%used in WL algorithm,
A more effective way
of controlling the schedule, $\alpha(t)$,
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t + 1},
  \label{eq:alpha_invt}
\end{equation}
%
where $t + 1$,
referred to as the ``time'' below,
is the number of updating steps
from the beginning of the simulation\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014}.
%
An intuitive explanation of this result\cite{
  marsili2006, barducci2008}
comes from the interpretation of
the bias potential %under the WL updating scheme, Eq. \eqref{eq:wl_update},
as a runtime average computed from all data points collected so far,
so that the addition of a new data point only
affects the average by its due weight,
which is the inverse of the current sample size, $t + 1$
(cf. Appendix \ref{sec:equilerr}).
%
Alternatively, we may derive it
from optimally planning the simulation lengths in the stages
and the reduction factor in the WL setup (cf. Appendix \ref{sec:wltoinvt}).

@@

In this section,
we will derive the optimal $\alpha(t)$
for the single-bin scheme,
Eq. \eqref{eq:wl_update},
as a preparation
for the more general multiple-bin scheme.
%
We will suggest a slight modification of Eq. \eqref{eq:alpha_invt}
whereby the origin of $t$ is shifted by an amount
determined from the initial error.
%
We will first express the error of
the bias potential, $V_i(t)$,
as a functional of $\alpha(t)$,
and then minimize it by variation.
%

\subsubsection{Error definition}

First, we define the net error in the bias potential
by deducting two contributions.
%
The first contribution comes from the target value,
Eq. \eqref{eq:Vi_target}.
%
For this purpose,
we introduce a shifted bias potential
%
\begin{equation}
  v_i(t)
  \equiv
  V_i(t)
  -
  \ln \frac { p^*_i }
            { p_i }
  ,
  \label{eq:v_def}
\end{equation}
%
which should tend to a constant of $i$
upon convergence.
Note that the updates to $v_i$'s are
equivalent to the updates to $V_i$'s,
since the shift
remains a constant during the course of simulation,
%%
%In terms of $v_i$'s, Eq. \eqref{eq:pi_p_V}
%becomes
%%
%\begin{equation}
%  \pi_i
%  =
%  \frac{                p_i \, e^{-v_i} }
%       { \sum_{j = 1}^n p_j \, e^{-v_j} }
%  \propto
%  p_i \, e^{-v_i}.
%  \label{eq:pi_p_v}
%\end{equation}
%%
%According to Eq. \eqref{eq:pi_p_V},

Second, the bias potential following Eq. \eqref{eq:wl_update}
generally grows over time.
%
Since a uniform increase of $V_i(t)$, or, equivalently, that of $v_i(t)$,
does not affect the
resulting distribution, $\pi_i(t)$,
%
the real deviation is more readily defined
from the difference between $v_i(t)$ and a baseline value, $\bar v(t)$,
that represents the average growth\cite{
dama2014},
%
\begin{equation}
  x_i(t) \equiv v_i(t) - \bar v(t)
  %=
  %v_i - \sum_{j = 1}^n \rho_j \, v_j
  ,
\label{eq:x_def}
\end{equation}
%
where
%
\begin{equation}
  \bar v(t)
  = \sum_{i = 1}^n \rho_i \, v_i(t)
  ,
\label{eq:vbar_def}
\end{equation}
for some distribution,
$\pmb \rho = (\rho_1, \dots, \rho_n)$
(with $\rho_i \ge 0$ and $\sum_{i = 1}^n \rho_i = 1$).
%
In this way,
the variable, $x_i(t)$,
represents the actual deviation of the bias potential, $V_i(t)$,
and the total error is given by
%
\begin{equation}
  \Err(t)
  =
  \sum_{i = 1}^n \rho_i \, \left\langle x_i^2(t) \right\rangle
  .
\label{eq:error_def}
\end{equation}

Although the above error weighting distribution, $\pmb\rho$,
is usually flat,
i.e. $\rho_i \equiv 1/n$,
in practice,
a non-flat distribution
is sometimes more convenient in analyzing
some multiple-bin updating schemes later on.
%
We also note that $\pmb\rho$
is independent of
the target distribution,
$\mathbf p = (p_1, \dots, p_n)$
(e.g. in well-tempered metadynamics,
$\mathbf p$ is an attenuated version of
$\mathbf p^*$\cite{
  barducci2008, dama2014}:
$\ln p_i = \ln {p^*_i}/(1+\Delta) + \mathrm{const.}$ for some positive $\Delta$,
whereas $\pmb\rho$ is flat).
%
From Eqs. \eqref{eq:x_def} and \eqref{eq:vbar_def},
we have
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, x_i(t) = 0
  .
\label{eq:rhox_sum}
\end{equation}
%

@@

To proceed, we will
approximate the finite difference Eq. \eqref{eq:wl_update}
by a differential equation
%
\begin{equation}
  \dot V_i(t)
  \equiv
  \frac{ dV_i(t) } { dt }
  =
  \alpha(t) \, \frac{ h_i(t) } { p_i },
  \label{eq:vt_diffeq}
\end{equation}
%
with
%
$
h_i(t) = \delta_{i, i(t)}
$
%\begin{equation}
%  h_i(t) = \delta_{i, i(t)}
%  ,
%  \label{eq:h_def}
%\end{equation}
%
being the instantaneous histogram,
which is equal to $1.0$
for the bin $i(t)$ at time $t$
or zero otherwise.
%
The shifted bias potential, $v_i(t)$,
follows the same differential equation
as the shift is a constant.

Next, we split $h_i(t)$ into a deterministic averaged part
and a random fluctuation part, $\zeta_i(t)$,
%
\begin{equation}
  h_i(t) = \langle h_i(t) \rangle + \zeta_i(t).
  \label{eq:h_split}
\end{equation}
%
The deterministic part can be related
to an ``ensemble average'',
where the ensemble consists of many similar simulation copies
sharing the same schedule $\alpha(t)$
and the same bias potential at time $t$.
%
The initial states and the stochastic forces
during the process may differ, however.

@@

The total error
at the end of the simulation, $t = T$,
as defined in Eq. \eqref{eq:error_def},
now depends implicitly on the function $\alpha(t)$
via $x_i(T)$.
%
Our aim is to find the $\alpha(t)$
that minimizes this error.
%%
%It suffices to consider
%a one-variable problem that
%minimizes each individual
%$\left\langle x_i^2(T) \right\rangle$.

%\subsubsection{\label{sec:onevar}
%Optimization of a one-variable problem}

\subsubsection{Optimal schedule for the single-bin scheme}

>>>>

The schedule of reducing
the updating magnitude,
therefore, critically affects
the precision of the final bias potential,
and hence that of the PMF\cite{laio2005, bussi2006, poulain2006,
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007, min2007,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014,
barducci2008, dickson2011, dama2014}.
%
For the single-bin case, the optimal updating schedule
is a function of the
inverse time\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007,
morozov2007, zhou2008}
or
the inverse number of updating steps.
\note{
  About the origin of the inverse-time formula.
  Besides the Belardinelli paper,
  we can also find it in Eq. (6) of reference \cite{liang2007},
  if I am not too mistaken.
  But I find the paper is a bit hard to read.
  There, however, the formula is $t_0/t$ instead of $1/t$.
}
The same schedule also applies to a variant of
the WL algorithm\cite{langfeld2012, pellegrini2014}.
%
%The optimal schedule for metadynamics,
%however, is less studied.

In this study,
we present a method of computing
the optimal schedule
for an adaptive FDS method
of a general updating scheme,
including the single-bin and Gaussian one
used in the WL algorithm and metadynamics.
%
Our method maps the optimization problem to a mechanical one,
in which the schedule plays the role of the velocity of
a free particle with a position-dependent mass,
and the final error becomes the action.
%
In this way, the optimal schedule
can be deduced from the equation of motion
of the particle that minimizes the action.
%
The resulting optimal schedule
depends on the updating window function
through the mass function.
%
While the optimal schedule in the single-bin case
is given by the known inverse-time formula,
that for Gaussian case is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
%In either case, however,
Ideally,
the optimal initial updating magnitude
can be set to
half of the previous equilibrium value,
which amounts to a shift of the origin of time
in the single-bin case.
%
Further, we will show that
the single-bin updating scheme
is optimal for asymptotic convergence,
and it can be generalized
to a class of asymptotically optimal
bandpass updating schemes.
%
The latter, however,
allow one to selectively flatten only
a few long-range histogram modes,
and to avoid introducing local noise
into the bias potential during the updating process.
%
The optimal schedule of bandpass updating schemes
is also the above origin-shifted inverse-time schedule.
%
Interestingly, several previous FDS algorithms\cite{
  langfeld2012, pellegrini2014,
  neuhaus2006, *neuhaus2007}
can be cast as special cases of bandpass updating schemes.

@@

Consider the problem of computing
the distribution, $p^*_i$,
along a discrete quantity, $i$,
for a given system.
%
%For example, $i$ can be the energy, $E$,
%in a lattice spin model or the temperature index
%in a simulated tempering simulation\cite{
%marinari1992, *lyubartsev1992}.
%
For a continuous quantity, $z$,
%which is often the case in a (classical) molecular system,
we can discretize it
such that the integer $i$ represents
the index of a small interval, or a bin,
$(z, z + \Delta z)$.
%
%The distribution is normalized as
%$\sum_{i = 1}^n p^*_i = 1$.

%For a large system at temperatures of interest,
%the distribution, $p^*_i$, often tends to
%be localized,
%%
%and to explore the global shape
%of the PMF, $-\ln p^*_i$,
%it is often advantageous to perform
%biased sampling that targets
%a wider and smoother distribution, $p_i$.
%%
%%Here, we refer to simulations that target
%%a flat or nearly flat distribution
%%as entropic or multicanonical sampling.
%
%To do so, we introduce a bias potential, $V_i$,
The challenge of sampling a multimodal distribution, $p^*_i$,
with deep valley(s) can often be eased by
biased sampling targeting a smoother distribution, $p_i$.
%
%
%where the proportionality constant, $C_i$,
%is found from the normalization condition,
%$\sum_{i = 1}^n \pi_i = 1$,
%in the second step.
%
%In particular,
%to make $\pi_i$ flat,
%the bias potential $V_i$
%must coincide with $\ln p^*_i$,
%up to an additive constant.
%%
%In practice, $V_i$ may be non-optimal or contain error,
%and thus $\pi_i$ need not
%coincide with the intended distribution $p_i$.
%
We will broadly define
a flat-distribution-sampling (FDS) simulation
as a biased simulation
targeting a flat $p_i$ (i.e. $p_i \equiv 1/n$)
or nearly flat one\cite{
dayal2004, *trebst2004, barducci2008, singh2011}.
%
By introducing a bias potential, $V_i$,
the biased equilibrium distribution becomes
%
\begin{equation}
  \pi_i
  %=
  %C_i \, p^*_i \, e^{-V_i}
  =
  \frac{                p^*_i \, e^{-V_i} }
       { \sum_{j = 1}^n p^*_j \, e^{-V_j} }
  \propto
  p^*_i \, e^{-V_i}
  ,
  \label{eq:pi_p_V}
\end{equation}
%
and our aim is to find a biased potential
such that the resulting $\pi_i$ would coincide
with the desired $p_i$.
%
Upon convergence, the bias potential would satisfy
%
\begin{equation}
  V_i \to \ln p^*_i - \ln p_i + \mathrm{const.}
  ,
  \label{eq:Vi_target}
\end{equation}
%
%according to Eq. \eqref{eq:pi_p_V}.
%which allows us to deduce the PMF, $-\ln p^*_i$,
%directly from the bias potential.
%
and the PMF, $-\ln p^*_i$, can be readily 
deduced from the bias potential. % $V_i$.
\note{
  For example, in the context of simulated tempering (Zhang and Ma, 2007),
  $p^*_i$ is replaced by $\ln Z_i$,
  $p_i$ is the weight $\zeta_i$,
  and the bias potential, $V_i$, is $\ln (\tilde Z_i / \zeta_i)$.
  Upon convergence, we have $\ln \tilde Z_i \to \ln Z_i$
  up to a constant shift.
}

Many methods\cite{mezei1987, berg1992, *lee1993,
wang2001, *wang2001pre, huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012}
have been developed to find the proper bias potential.
%
In a class of equilibrium FDS methods\cite{
  mezei1987, berg1992, *lee1993, marinari1992, *lyubartsev1992},
the bias potential, $V_i$, is
estimated a priori and fixed
during simulation,
%
and a correction to the bias potential
is derived from the normalized histogram, $H_i$,
accumulated from the simulation as
%
%Using $H_i$ as an estimate of $\pi_i$ in
%Eq. \eqref{eq:pi_p_V}
%allows us to correct $V_i$ as
%
\begin{equation}
  V^\mathrm{corr}_i
  =
  V_i
  +
  \ln \frac{ H_i }
           { p_i }.
  \label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    H_i \approx \pi_i
    &\propto p^*_i \, e^{-V_i},
    \\
    p_i
    &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
However, this formula often requires a long
accumulation period for a precise histogram,
and thus disallows
continuous improvement of the bias potential
%(hence that of the PMF)
as simulation lengthens.

An other class of adaptive FDS methods,
including the WL algorithm\cite{wang2001, *wang2001pre}
and metadynamics\cite{huber1994, laio2002, *laio2008, *barducci2011, *sutto2012},
address the above shortcoming
by continuously and incrementally updating the bias potential,
and we will focus on these methods below.
%%
%%These methods incrementally update the bias potential
%%and can approximately sample the target distribution.
%%
%However,
%the on-the-fly updates break the microscopic reversibility
%of the underlying sampling mechanism.
%Further, failure to properly reduce the magnitude of
%the histogram-based increments
%would lead to roughness or errors in the final estimate of the PMF.
%%
%%Thus, for a conservative practitioner,
%%the E-FDS methods appear more appealing
%%once a sufficiently accurate bias potential is available.
%%
%To minimize this effect,
%many adaptive methods adopt a schedule to
%reduce the updating magnitude
%over the course of simulation\cite{
%marsili2006,
%belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
%liang2007,
%barducci2008}.
%%
%%Thus, in late stages of a long simulation,
%%the updating becomes negligible and
%%the adaptive simulation tends to
%%the corresponding equilibrium one.
%%
%%if the bias potential were converged to the exact value,
%%this adaptive FDS tends to the ideal equilibrium counterpart,
%%particular equilibrium simulation can be thought as
%%the optimal asymptotic limit of the adaptive counterpart,
%%
%For example, one can show that
%if the magnitude is controlled by
%the optimal inverse-time prescription\cite{
%belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
%liang2007,
%morozov2007, zhou2008,
%komura2012, *caparica2012, *caparica2014},
%the WL algorithm is equivalent to the ideal equilibrium counterpart
%in improving the precision of the bias potential,
%thus allowing for maximum efficiency
%(cf. Appendix \ref{sec:equilerr}).
%%the bias potential in the WL algorithm
%%will evolve similar to Eq. \eqref{eq:vcorr_equil}
%%with the precision of the histogram as simulation lengthens.

@@

%We will mainly focus on the adaptive methods below.
%
In the original WL algorithm\cite{wang2001, *wang2001pre},
the bias potential, $V_i(t)$, is updated
at each sampling step $t$ as
%
\begin{equation}
  V_i(t+1)
  =
  V_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { p_{i(t)} }
  ,
\label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
$\delta_{i, \, i(t)}$ is the Kronecker delta function,
which is $1$ when $i = i(t)$ or $0$ otherwise,
and $\alpha(t)$ is the updating magnitude.
%
We refer to this scheme as the \emph{single-bin scheme},
for it applies only to the bias potential
at the current bin, $i(t)$.
%
In a more general \emph{multiple-bin scheme},
including the Gaussian updating scheme
used in metadynamics,
several neighboring bins are updated:
%
\begin{equation}
  V_i(t+1)
  =
  V_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) }
       { p_{i(t)} }
  .
  \label{eq:mbin_update}
\end{equation}
%
The Gaussian updating scheme used in metadynamics
is an example, with $w_{ij}$ being a Gaussian of $i-j$.
%
Note that if the updating occurs
only every few sampling steps,
then $t$ means the number of updating steps.

@@

%For the single-bin scheme
%used in WL algorithm,
A more effective way
of controlling the schedule, $\alpha(t)$,
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of updating steps
from the beginning of the simulation\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014}.
%
This can be derived from a general result\cite{
  robbins1951, pellegrini2014}
and an intuitive explanation\cite{
  marsili2006, barducci2008}
comes from the interpretation of
the bias potential %under the WL updating scheme, Eq. \eqref{eq:wl_update},
as a runtime average computed from all data points collected so far,
so that the addition of a new data point only
affects the average by its due weight,
the inverse of the current sample size, $t$
(cf. Appendix \ref{sec:equilerr}).
%
We may also derive it
from optimally planning the simulation lengths in the stages
and the reduction factor in the WL setup (cf. Appendix \ref{sec:wltoinvt}).

@@

In this section,
we will derive the optimal $\alpha(t)$
for the single-bin scheme,
Eq. \eqref{eq:wl_update},
as a preparation
for the more general multiple-bin scheme.
%
%We will suggest a slight modification of Eq. \eqref{eq:alpha_invt}
%whereby the origin of $t$ is shifted by an amount
%determined from the initial error.
%
We will first express the error of
the bias potential, $V_i(t)$,
as a functional of $\alpha(t)$,
and then minimize it by variation.
%

We define the net error in the bias potential
by deducting two contributions.
%
First, to deduct the target value given by
Eq. \eqref{eq:Vi_target},
we introduce a shifted bias potential
%
\begin{equation}
  v_i(t)
  \equiv
  V_i(t)
  -
  \ln \frac { p^*_i }
            { p_i }
  ,
  \label{eq:v_def}
\end{equation}
%
which should tend to a constant of $i$
upon convergence.
Note that the updates to $v_i(t)$ are
equivalent to the updates to $V_i(t)$,
since the shift
remains a constant during the course of simulation,
%%
%In terms of $v_i$'s, Eq. \eqref{eq:pi_p_V}
%becomes
%%
%\begin{equation}
%  \pi_i
%  =
%  \frac{                p_i \, e^{-v_i} }
%       { \sum_{j = 1}^n p_j \, e^{-v_j} }
%  \propto
%  p_i \, e^{-v_i}.
%  \label{eq:pi_p_v}
%\end{equation}
%%
%According to Eq. \eqref{eq:pi_p_V},

Second, the bias potential following Eq. \eqref{eq:wl_update}
generally grows over time.
%
Since a uniform increase of $V_i(t)$, or, equivalently, that of $v_i(t)$,
does not affect the
resulting distribution, $\pi_i(t)$,
%
the real deviation, $x_i(t)$, of the bias potential
depends only on the difference between $v_i(t)$
and a baseline value, $\bar v(t)$,
for average growth\cite{
dama2014},
%
\begin{equation}
  x_i(t) \equiv v_i(t) - \bar v(t)
  %=
  %v_i - \sum_{j = 1}^n \rho_j \, v_j
  ,
\label{eq:x_def}
\end{equation}
%
where
%
\begin{equation}
  \bar v(t)
  = \sum_{i = 1}^n \rho_i \, v_i(t)
  ,
\label{eq:vbar_def}
\end{equation}
for some distribution,
$\pmb \rho = (\rho_1, \dots, \rho_n)$
(with $\rho_i \ge 0$ and $\sum_{i = 1}^n \rho_i = 1$).
%
The total error is given by
%
\begin{equation}
  \Err(t)
  =
  \sum_{i = 1}^n \rho_i \, \left\langle x_i^2(t) \right\rangle
  .
\label{eq:error_def}
\end{equation}

%Although the above error weighting distribution, $\pmb\rho$,
%is usually flat,
%i.e. $\rho_i \equiv 1/n$,
%in practice,
%a non-flat distribution
%is sometimes more convenient in analyzing
%some multiple-bin updating schemes later on.
%
Note that the distribution, $\pmb\rho$, is arbitrary
and independent of
the target distribution,
$\mathbf p = (p_1, \dots, p_n)$,
although it is usually flat.\footnote{In
well-tempered metadynamics,
$\mathbf p$ is an attenuated version of
$\mathbf p^*$\cite{
  barducci2008, dama2014}:
$\ln p_i = \ln {p^*_i}/(1+\Delta) + \mathrm{const.}$ for some positive $\Delta$,
whereas $\pmb\rho$ is flat.}
%
From Eqs. \eqref{eq:x_def} and \eqref{eq:vbar_def},
we have
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, x_i(t) = 0
  .
\label{eq:rhox_sum}
\end{equation}
%

@@

To proceed, we will
approximate the finite difference Eq. \eqref{eq:wl_update}
by a differential equation\footnote{This
approximation requires us to shift the origin of time by $1$ in some cases.}
%
\begin{equation}
  \dot V_i(t)
  \equiv
  \frac{ dV_i(t) } { dt }
  =
  \alpha(t) \, \frac{ h_i(t) } { p_i },
  \label{eq:vt_diffeq}
\end{equation}
%
with
%
$
h_i(t) = \delta_{i, i(t)}
$
%\begin{equation}
%  h_i(t) = \delta_{i, i(t)}
%  ,
%  \label{eq:h_def}
%\end{equation}
%
being the instantaneous histogram,
which is equal to $1.0$
for the bin $i(t)$ at time $t$
or zero otherwise.
%
The shifted bias potential, $v_i(t)$,
follows the same differential equation
as the shift is a constant.

We can split $h_i(t)$ into a deterministic averaged part
and a random fluctuation part, $\zeta_i(t)$,
%
\begin{equation}
  h_i(t) = \langle h_i(t) \rangle + \zeta_i(t).
  \label{eq:h_split}
\end{equation}
%
The deterministic part can be related
to an ``ensemble average'',
where the ensemble consists of many similar simulation copies
sharing the same schedule $\alpha(t)$
and the same bias potential at time $t$.
%
The initial states and the stochastic forces
during the process may differ, however.

@@

The formal solution of Eq. \eqref{eq:dxdt_singlebin} is
%
\begin{equation}
x_i(T)
=
x_i(0) \, e^{-q(T)}
+
\int_0^T
  \dot u\bigl( q(t) \bigr) \, \xi_i(t) \, dt,
\notag
%\label{eq:xt_solution}
\end{equation}
%
where
%
$
q(t) \equiv \int_0^t \alpha(t') \, dt',
$
%
and
%
\begin{align}
u(q')
&\equiv
e^{q' - q(T)}.
\label{eq:u_def}
\end{align}

\subsubsection{Optimal schedule}

At the end of period $T$,
the total error defined in Eq. \eqref{eq:error_def}
is given by
%
\begin{align}
  E(T)
  =
  E(0) \, e^{-2 \, q(T)}
  +
  \Gamma \,
  \int_0^T
    {\dot u}^2\bigl( q(t) \bigr) \,
    \, dt
  ,
  \label{eq:ET_average}
\end{align}
where
\begin{equation}
  \Gamma \equiv \sum_{i=1}^n \rho_i \, G_i.
  \label{eq:Gamma_def}
\end{equation}
